---
title: "S. Korea City/Disease Correlation"
author: "Annika Gaglani"
date: "2023-10-29"
output: html_document
---

#install packages and read them into library
```{r}
install.packages("tidyverse")
install.packages("modelr")
install.packages("dplyr")
install.packages("ggplot2")
library(tidyverse)
library(modelr)
library(dplyr)
library(ggplot2)
options(na.action = na.warn)
```

#import data and clean up
```{r}
#CO
co_data <- read.csv("co.csv")
co_data
#SOx
sox_data <- read.csv("sox.csv")
sox_data
```
```{r}
#population by city data
pop_data <- read.csv("P-ai Population by City Data - cleaned.csv")
pop_data
```

#joins for Incheon
```{r}
#CO and population for Incheon
incheon_co_pop_join <- left_join(co_data, pop_data, by = "X") %>% rename("Incheon_pol" = Incheon.x, "Incheon_pop" = Incheon.y)
incheon_co_pop_join
```

```{r}
#SOx and pollution for Incheon
incheon_sox_pop_join <- left_join(sox_data, pop_data, by = "X") %>% rename("Incheon_pol" = Incheon.x, "Incheon_pop" = Incheon.y)
incheon_sox_pop_join
```

#joins for Seoul
```{r}
#CO and population for Seoul
seoul_co_pop_join <- left_join(co_data, pop_data, by = "X") %>% rename("Seoul_pol" = Seoul.x, "Seoul_pop" = Seoul.y)
seoul_co_pop_join
```

```{r}
#SOx and populationf for Seoul
seoul_sox_pop_join <- left_join(sox_data, pop_data, by = "X") %>% rename("Seoul_pol" = Seoul.x, "Seoul_pop" = Seoul.y)
seoul_sox_pop_join
```


#models for Incheon
```{r}
#CO
incheon_co_mod <- lm(Incheon_pol ~ Incheon_pop, data = incheon_co_pop_join)
incheon_co_mod

incheon_co_pop_join %>%
  ggplot() +
  geom_col(aes(x = Incheon_pop, y = Incheon_pol)) #+
  #geom_abline(aes(intercept = incheon_co_mod$coefficients[1],
                  #slope = incheon_co_mod$coefficients[2]), color = "red")
```
Not showing much of a correlation between CO and population (lots of fluctuation), thinking that COVID might be a confounding variable.

```{r}
#SOx
incheon_sox_mod <- lm(Incheon_pol ~ Incheon_pop, data = incheon_sox_pop_join)
incheon_sox_mod

incheon_sox_pop_join %>%
  ggplot() +
  geom_col(aes(x = Incheon_pop, y = Incheon_pol)) #+
  #geom_abline(aes(intercept = incheon_sox_mod$coefficients[1],
                  #slope = incheon_sox_mod$coefficients[2]), color = "red")
```
Tried changing the bin width on these plots and it didn't make anything more clear.

Again not showing much of a correlation for SOx. Worth noting that there is a outlier on the high side in both CO and SOx around the same time period. Also viewing as a bar graph instead of a scatterplot shows that there is a slight correlation between SOx and population - Incheon's population has been increasing so there has been a slight increase in pollution. 

#models for Seoul
```{r}
#CO
seoul_co_mod <- lm(Seoul_pol ~ Seoul_pop, data = seoul_co_pop_join)
seoul_co_mod

seoul_co_pop_join %>%
  ggplot() +
  geom_col(aes(x = Seoul_pop, y = Seoul_pol)) #+
  #geom_abline(aes(intercept = seoul_co_mod$coefficients[1],
                  #slope = seoul_co_mod$coefficients[2]), color = "red")
```
There's a lot of fluctuation but also showing a slight increase trend. 

```{r}
#SOx
seoul_sox_mod <- lm(Seoul_pol ~ Seoul_pop, data = seoul_sox_pop_join)
seoul_sox_mod

seoul_sox_pop_join %>%
  ggplot() +
  geom_col(aes(x = Seoul_pol, y = Seoul_pop))
```
Not showing much of a correlation either??
