---
title: "disease_analysis"
output: pdf_document
date: "2023-10-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(broom)
```

## R Markdown

```{r} 
#*
# read in and edit data sets 
# disease
disease <- read_csv("Downloads/disease.csv", 
    col_types = cols(...1 = col_skip(), `2020.01 num_patients` = col_double(), 
        len_hospital_stay...5 = col_double(), 
        num_insurance_claims...6 = col_double(), 
        total_insurance_cost...7 = col_double(), 
        outpocket_cost...8 = col_double(), 
        `2020.02 num_patients` = col_double(), 
        len_hospital_stay...10 = col_double(), 
        num_insurance_claims...11 = col_double(), 
        total_insurance_cost...12 = col_double(), 
        outpocket_cost...13 = col_double(), 
        region...184 = col_skip(), disease_code...185 = col_skip()))
# no2
no2 <- read_csv("Downloads/no2.csv", col_types = cols(Daegu = col_double(), 
    Gangwon = col_double(), Jeju = col_double()))
no2_cleaned <- na.omit(no2)
colnames(no2_cleaned)[1] <- "X"

# co
co <- read.csv("Downloads/co.csv")

# ozone, same as no2 ??? 
ozone <- read_csv("Downloads/ozone.csv", 
    col_types = cols(Daegu = col_double(), 
        Gangwon = col_double(), Jeju = col_double()), 
    skip = 1)
colnames(ozone)[1] <- "date"
ozone_cleaned <- na.omit(ozone)


# pm10
pm10 <- read_csv("Downloads/pm10.csv", col_types = cols(Daejeon = col_double(), 
    Gangwon = col_double()))
pm10_cleaned <- na.omit(pm10)

# sox
sox <- read.csv("Downloads/sox.csv")
```


```{r}
# find summary stats to see difference in diseases 
summary(disease)

# filter 
disease_J411 <- disease %>%
  filter(disease_code...2 == "J411")

# look at only number of patients then sort for region for each year 
disease_J411_2020 <- select(disease_J411, contains("region") | contains("2020"))
# compare air pollutants to each month to see seasonal and total overall to see if there is a rise in disease with air pollutant levels 

# transpose to compare to air pollutants 
disease_J411_2020_t <- as.data.frame(t(disease_J411_2020))
disease_J411_2020_t<-disease_J411_2020_t[-1,]
colnames(disease_J411_2020_t) <- disease_J411_2020$region...3
disease_J411_2020_t[] <- lapply(disease_J411_2020_t, as.numeric)
ozone_cleaned_2020 <- ozone_cleaned[(34:45),]

# test correlation for seasons between air pollutants and 2020 number of patients 


# spring 
cor.test(ozone_cleaned_2020$Seoul[3:5], disease_J411_2020_t$Seoul[3:5])

# summer 
cor.test(ozone_cleaned_2020$Seoul[6:8], disease_J411_2020_t$Seoul[6:8])

# fall 
cor.test(ozone_cleaned_2020$Seoul[9:11], disease_J411_2020_t$Seoul[9:11])

# winter 




```

no correlation between season and ozone



```{r}
#*
# correlation analysis all time vs all regions
cor(co$X, rowMeans(co[, -1]))
cor(sox$X, rowMeans(sox[, -1]))
cor(no2_cleaned$X, rowMeans(no2_cleaned[, -1]))
cor(pm10_cleaned$...1, rowMeans(pm10_cleaned[, -1]))


# gives you the correlation individually for each region instead of average ^
correlations <- lapply(co[, -1], function(x) cor(co$X, x))


# merge co, no2, sox to see if any relationship between them 
# correlation 
co_sox <- left_join(co, sox, by = join_by(X == X))

cor(co_sox$Gwangju.x, co_sox$Gwangju.y)

# look at seasonal correlation 
cor(co_sox$Busan.x[6:8], co_sox$Busan.y[6:8])
    
# linear regression 
lm_Seoul <- lm(co_sox$Seoul.x ~ co_sox$Seoul.y)
summary(lm_Seoul)
```
^ from these tests it seems there is no / not strong relationship between time and air pollutants for each of these regions 
although found a paper that has already analyzed air pollution and urban head island , found a positive relationship with interactions with them during seasons

CO and SOX have some relationship between them 
^ seasonal relationship, it varies for each region, so must specifically analyze that !!!!!!


```{r}
# ANOVA to see if there are stat sig diff in disease rates among regions , no there isnt i think

# aov(disease_code...2 ~ 2020.01 num_patients , data = disease)

# ANOVA to see if differences in diff groups among the time for diff disease codes 
# aov(, data = disease)

# checking num of patients between diseases 

# tukey HSD to check which regions are diff
```

- check if data is normal for ANOVA



# calculate disease incidence 
1. num patients for each disease code in each region and time period (we have)
2. calc population at risk for it 
3. incidence rate = number of cases / total pop at risk then * 100,000
seoul pop = 9.7 mil

